---
title: "01_basics"
author: "Fran Camacho"
date: "2025-03-25"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Intro

...


## 2. The basics


### 2.4.3 Vectors and data frames



```{r}
age <- c(10, 28, 48, 47, 71, 22, 80, 48, 30, 31)
age

age_months <- age * 12
age_months
```



```{r}
purchase <- c(14, 20, 59, 2, 12, 22, 160, 34, 34, 29)
purchase
```

It would be nice to combine these two vectors into a table -> dataframe

```{r}
bookstore <- data.frame(age, purchase)
bookstore
```

The fact that

```{r}
x <- 4
x
```


renders the output  [1] 4

tells us that **x in fact is a vector, albeit with a single element**. Almost everything in R is a vector, in one way or another.


Exercise 2.4 Do the following:

Create two vectors, height and weight, containing the heights and weights of five fictional people (i.e., just make up some numbers!).
Combine your two vectors into a data frame.

```{r}
names <- c("foo", "mola", "nandia", "fefe", "ogogo")
heights <- c(177, 220, 190, 168, 201)
weights <- c(88, 110, 95, 60, 90)

people1 <- data.frame(names, heights, weights)
people1
```


### 2.4.4 Functions



