---
title: "01_basics_dataframes"
author: "Fran Camacho"
date: "2025-05-13"
output: word_document
---

## 2.10 Data frames and data types

### 1 Types and structures

```{r}
x <- 6
y <- "Scotland"
z <- TRUE

class(x)
class(y)
class(z)
```

Vectors:

```{r}
numbers <- c(6, 9, 12)

class(numbers)
```

class returns the data type of the elements of the vector. 

So what happens if we put objects of different types together in a vector?

```{r}
all_together <- c(x, y, z)
all_together

class(all_together)
```

### 2 Types of tables


```{r}
# Bookstore example
age <- c(28, 48, 47, 71, 22, 80, 48, 30, 31)
purchase <- c(20, 59, 2, 12, 22, 160, 34, 34, 29)
bookstore <- data.frame(age, purchase)

str(bookstore)
bookstore
```

```{r}
# Animal sleep data
library(ggplot2)
View(msleep)
```

Here’s the rundown of four common types:

- matrix: a table where all columns must contain objects of the same type (e.g., all numeric or all character). 
Uses less memory than other types and allows for much faster computations, but it is difficult to use for certain types of data manipulation, plotting and analyses.
    
- data.frame: the most common type, where different columns can contain different types 
(e.g., one numeric column, one character column).

- data.table: an enhanced version of data.frame.
    
- tbl_df (“tibble”): another enhanced version of data.frame.



```{r}
# First, an example of data stored in a matrix:
?WorldPhones
class(WorldPhones)
View(WorldPhones)

# Next, an example of data stored in a data frame:
?airquality
class(airquality)
View(airquality)

# Finally, an example of data stored in a tibble:
library(ggplot2)
?msleep
class(msleep)
View(msleep)
```

 you can convert objects into other types:
 
```{r}
WorldPhonesDF <- as.data.frame(WorldPhones)
class(WorldPhonesDF)

airqualityMatrix <- as.matrix(airquality)
class(airqualityMatrix)
```
 

**Exercise 2.23** 

Matrix tables can be created from vectors using the function of the same name. 
Using the vector x <- 1:6 use matrix to create the following matrices:

(1 2 3
 4 5 6)

```{r}
x <- 1:6

matrix(x, nrow = 2, ncol = 3, byrow = TRUE)
```


and

(1 4
 2 5
 3 6)
 
```{r}
x <- 1:6

matrix(x, nrow = 3, ncol = 2, byrow = FALSE)
```
 

## 2.11 Vectors in data frames


```{r}
airquality_raw <- data.frame(airquality) 
```

```{r}
# Extract the Temp vector:
airquality$Temp
```

```{r}
# Compute the mean temperature:
mean(airquality$Temp)
```

```{r}
# First, we check the order of the columns:
names(airquality)
# We see that Temp is the 4th column.

airquality[5, 4]    # The 5th element from the 4th column,
                    # i.e. the same as airquality$Temp[5]

airquality[5,]      # The 5th row of the data
```


```{r}
airquality[, 4]     # The 4th column of the data, like airquality$Temp

airquality[[4]]     # The 4th column of the data, like airquality$Temp
```


```{r}
airquality[, c(2, 4, 6)] # The 2nd, 4th and 6th columns of the data

airquality[, -2]    # All columns except the 2nd one

airquality[, c("Temp", "Wind")] # The Temp and Wind columns
```










