---
title: "02_cornerstones_of_stadistics"
author: "Fran Camacho"
date: "2025-08-06"
output:
  html_document:
    df_print: paged
---

# 02 - Cornerstones of Stadistics


```{r}
#Libraries

#https://tidyverse.r-universe.dev/tidyverse
if (!require(tidyverse)) install.packages('tidyverse', repos = c('https://tidyverse.r-universe.dev', 'https://cloud.r-project.org'))
library(tidyverse)




```

## 3.1 The three cultures

There are three main schools in statistical modelling: 
the frequentist school, 
the Bayesian school, and 
the machine learning school.


## 3.2 Frequencies, proportions, and cross-tables


```{r}
library(palmerpenguins)
?penguins
```


- Frequency tables:
Frequency tables are used to summarise the distribution of a single variable. 

Without pipes:
```{r}
table(penguins$species)
```

With pipes:
```{r}
library(dplyr)
penguins |> select(species) |> table()
```


- Publication-ready tables

The table in the previous section is fine for data exploration but requires some formatting if you wish to put it in a report. 
If instead you want something publication-ready that you can put straight into a report or a presentation, 
I recommend using either the **gtsummary** package or the **ivo.table** package. 

```{r}
# Install the packages:
#install.packages(c("gtsummary", "flextable", "ivo.table"))

if (!require(gtsummary)) install.packages('gtsummary', dependencies = T)

if (!require(flextable)) install.packages('flextable', dependencies = T)
library(flextable)
if (!require(ivo.table)) install.packages('ivo.table', dependencies = T)
library(ivo.table)
```


Without pipes:
```{r}
library(gtsummary)

tbl_summary(penguins[,"species"])
```

With pipes:
```{r}
library(dplyr)
library(gtsummary)

penguins |>
  select(species) |>
  tbl_summary()
```


The **ivo.table** package and the **ivo_table** function offer a similar table constructed in the same way:

Without pipes:
```{r}
library(ivo.table)

ivo_table(penguins[,"species"])
```


With pipes:
```{r}
library(dplyr)
library(ivo.table)

penguins |>
  select(species) |>
  ivo_table()
```

You can modify the settings to change the colours and fonts used, and to show percentages instead of counts:
```{r}
penguins |>
  select(species) |>
  ivo_table(color = "darkred",
            font = "Garamond",
            percent_by = "row")
```

If you prefer, you can also get the table in a long format that is useful when your variable has many different levels:
```{r}
penguins |>
  select(species) |>
  ivo_table(long_table = TRUE)
```

Finally, the table can be exported to Word as follows:
```{r}
library(flextable)

penguins |>
  select(species) |>
  ivo_table() |> 
  save_as_docx(path = "my_table.docx")
```



- Contingency tables

Contingency tables, also known as cross-tabulations or cross-tables, are used to summarise the relationship between two or more variables.


