---
title: "deviation.Rmd"
author: "Fran Camacho"
date: "2025-03-28"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## THE MEAN DEVIATION


**Example 5.3.1**. The weights, in grams, of 10 pieces of scrap metal from a metal fab-
rication process are 2.63, 2.68, 2.57, 2.73, 2.56, 2.62, 2.38, 2.45, 2.35, 2.53. Compute
the mean deviation about the mean and mean deviation about the median.

```{r}
measures <- c(2.63, 2.68, 2.57, 2.73, 2.56, 2.62, 2.38, 2.45, 2.35, 2.53)

paste("Mean  : ",mean(measures))
paste("Median: ",median(measures))
```



```{r}
dif_mean <- abs(measures - mean(measures))
dif_mean
dif_median <- abs(measures - median(measures))
dif_median
n <- length(measures)

mean_abs_dev <- sum(dif_mean)/n
median_abs_dev <- sum(dif_median)/n

paste("Mean-absolute deviation: ", mean_abs_dev)
paste("Median-absolute deviation: ", median_abs_dev)
```

"mad" returns a different value

```{r}
# Mean-absolute deviation (MAD)
# https://search.r-project.org/CRAN/refmans/ie2misc/html/madstat.html

paste("Mean-absolute deviation: ", mad(measures))
```

?????????????????????????????????????????????!!!!!!!!!!!!!!!


**First and third quartile**

```{r}
quantile(measures, 0.25) # first quartile
quantile(measures, 0.75) # third quartile
```


**Other quantiles**

Any quantile can be computed with the quantile() function. For instance, the 4th decile or the 98th percentile:

```{r}
quantile(measures, 0.4)  # 4th decile
quantile(measures, 0.98) # 98th percentile
```


**Interquartile range**

The interquartile range (i.e., the difference between the first and third quartile) can be computed with the IQR() function:

```{r}
IQR(measures)
```

The interquartile range measures the spread bounding the middle 50 percent of the values of the data.


## THE VARIANCE AND THE STANDARD DEVIATION


- The variance of a set of observations x1 , . . . , xn is defined as the arithmetic mean
of the squares of the deviations of the individual observations from their arithmetic mean

- The positive square root of the variance is called the standard deviation.


```{r}

paste("Variance: ", var(measures))
paste("Standard. deviation: ", sd(measures))

```

The standard deviation of a set of observations will be always positive even if all the observations in the data are negative. 
Standard deviation provides a summary of variability or spread or scatter which reflects how scattered are the measurements from
their arithmetic mean.


**Example 5.4.1**. The heights of five people in centimeters are 183, 163, 172, 157, and 169. 
Find the standard deviation of this data.


```{r}
heights <- c(183, 163, 172, 157, 169)
var(heights)
sd(heights)
```

In the book where I took this exercises from, it says var. is 76.96, and std. dev. is 8.77 !!!!!!!!!!!!!!!!!


## SKEWNESS IN THE DATA

If the observations in a data are spread in a perfect symmetrical manner about an average, then the mean and median are equal. 
If the data has a single mode and it is perfectly symmetric, then the mean,the median and the mode coincide.

Skewness is zero when the data is symmetric.

If the frequencies and class intervals on the right of the median is an exact replica of those on the left of the median then the data is symmetric, 
if not the data is asymmetric. 
Asymmetry may happen when on one side the data is spread on a wider range so that the arithmetic mean of the whole data may lie on to the right or
to the left of the median. If the arithmetic mean is shifted to the right of the median, as a result wider spread with smaller frequencies, 
then we say that the data is skewed to the right or positively skewed. If the same thing happens on the left of the median then
we say that the data is skewed to the left or negatively skewed.



